---
title: "Data Wrangling"
author: "Scout Leonard"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
```

## Warm Up 

```{r}
income_data_wider <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable,
              values_from = estimate)

income_data_wider
```

# Practice Some Tidyverse Functions for Data Wrangling 

- `filter()`
- `select()`
- `mutate()`
- `case_when()`
- `group_by() %>% summarize()`
- piping things together in sequence 

### dplyr::filter() to keep of exclude rows based on out conditions

1.) Exact match (signle condition)

Only keep observations where the island is Biscoe

```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe") 
# check to see that we only have Biscoe with `unique()`
unique(ex_1$island)
```
2.) "and" statement within filter

Only keep observations if the penguins species is Chinstrap AND the year is 2008

```{r}
ex_2 <- penguins %>% 
  filter(species == "Chinstrap", year == "2008")

ex_2 <- penguins %>% 
  filter(species == "Chinstrap" & year == "2008")

ex_2 <- penguins %>% 
  filter(species == "Chinstrap") %>% 
  filter(year == "2008")
```

3.) Exclude Gentoos 
Keep observations if the penguins species is NOT "GEntoo"

```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```

Exercise

Create a subset forom penguins that only contains observations for Adelie penguins with a flipper length greater than or equal to 185mm. 

```{r}
ex_4 <- penguins %>% 
  filter(species == "Adelie") %>% 
  filter(flipper_length_mm >= 185)

unique(ex_4$species)

min(ex_4$flipper_length_mm)
```
4.) The %in% operator

Use the %in% operator to look for *any* matches within a vector, and return TRUE if the value does exist for our row. 

We want to keep observations where the island is Dream or Torgerson

```{r}
ex_5 <- penguins %>% 
  filter(island %in% c("Dream", "Torgerson"))

#if we use a double equals with a vector, the code looks for that vector in order, and so would skip some of the data we want in this example! Add a ! before the variable to exclude a vector
```

### select() to select of exclude COLUMNS

5.) Keep a single column by name 

Only want to keep the species column 

```{r}
ex_6 <- penguins %>% select(species) %>% 
  pull(species)

ex_6

#Turn this to a vector instead of a dataframe  with pull()

#Rewrite this without the pipe operator

ex_6 <- select(.data = penguins, species)
```

6.) Keep multiple functions by name 

Create a subset that contains species, bill_length_mm thorugh flipper_length-mm, and year

```{r}
ex_7 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)
```

7.) Choose or exclude columns based on position 

```{r}
#excludes 4th column 
ex_8 <- penguins %>% 
  select(-4)
```

8.) A piped sequence

Write a piped sequence where you first only keep columns for species, body_mass_g, and island , and second: filter to include only observations where bodymass is great than 3600

```{r}
ex_9 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)

#check filter
min(ex_9$body_mass_g)
```

### mutate() to add or update existing columns while keeping the ones already there

9.) Add a column that has body mass converted to kilograms 

```{r}
ex_10 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) %>% 
  mutate(flipper_length_m = flipper_length_mm / 1000)

#you can do multiple mutate statements in one mutate function 
```

10.) Mutate to update and existing column

```{r}
ex_11 <- penguins %>% 
  mutate(island_hooray = str_replace(island, pattern = "sen", replacement = "HOORAY!"))

# can be used to update spelling as an example. replaces "sen" anywhere in the column island with "HOORAY!" Update the column name with something different to add a new column and not mess up the original! 
```

11.) Update the class of the bill depth column to character

```{r}
ex_12 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))

class(ex_12$bill_depth_mm)
```

# Example 12: case_when(): your favorite if-else

Create a new column that contains "small" if penguinflipper length less than 185 mm, "medium" if otherwise
```{r}
ex_13 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
  ))
```

Task: Starting from penguins, create a new column called "bird_bill_bin" that contains: 

- if bill length < 38 mm : "short bill"
- if bill length >= 38 mm & < 43mm : "medium bill" 
- if bill length >= 43: : "long bill" 

```{r}
task_2 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 143 ~ "medium bill",
    bill_length_mm >= 143 ~ "long bill",
    TRUE ~ "long bill"
  ))

# last line says that if you have any other conditions are classified as long bill (this will claffify NAs as long bill!)
```

## group_by() %>% summarize() to find summary vlaues by group, and return them in a nice summary table

```{r}
mean(penguins$bill_depth_mm, na.rm = TRUE)
```

13.) Use group_by() to say there are groups within the speices column 

```{r}

#group_by() doesn't actually change the structure or content of your dataframe
ex_14 <- penguins %>% 
  group_by(species, island, year) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE), sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))

#makes a new summary table for you! 

ex_14
```

## Practice Task 

In a single pipe sequence, do the following: 

- Only keep penguins that have a body mass >3500 g
- Only keep columns for species, flipper_length_mm and sex
-  Create a summary table that contains the mean, sd, max, and min value of flipper length by species and sex (as the grouping variables)

```{r}
task_3 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE), 
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE), 
            max_flipper_length = max(flipper_length_mm, na.rm = TRUE), 
            min_flipper_length = min(flipper_length_mm, na.rm = TRUE)) %>% 
  drop_na(sex)

#for any penguins where the sex wasn't recorded, are not part of the summary table in drop_na(). you can also filter(!= na) for sex
```




