---
title: "Tidying Data"
author: "Scout Leonard"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
```

## Read in the wb_indicators.csv file, make sure to specify `na.rm = c("..,")`

```{r}
wb_indicators <- read_csv(here("data", "wb_indicators.csv"), na = c("..",""))
```

## Reshape so that year is a single column 

```{r}
wb_indicators_long <- wb_indicators %>% 
  pivot_longer(cols = `2001 [YR2001]`:`2020 [YR2020]`,
               names_to = "year",
               values_to = "indicator_value")
```

There are no rules about the bumber of things you can do in a pipe sequence. We can do this next part in the code above, but we are breaking it up to understand outputs. 

## Clean up the year column
The current class of the year column is character, but we want it to be numeric. 

```{r}
wb_data_sep <- wb_indicators_long %>% 
  separate(col = year, into = c("year", "year_chr"), sep =  " ") %>% 
  select(-c(year_chr, 'Country Code', 'Series Code'))
```

```{r}
df <- tribble(
  ~name, ~A, ~B, ~C,
  "teddy", 4, 5, 6,
  "khora", 9, 10, 11
)

df

# If A, B, C are different dog food types, convert this into tidy format

df_long <- df %>% 
  pivot_longer(cols = A:C, 
               names_to = "dog_food_type",
               values_to = "rankings")
```

You can think of a tibble as a dataframe. 

# Pivot wider to get all indicators in their own column

```{r}
wb_data_tidy <- wb_data_sep %>% 
  drop_na('Series Name') %>% 
  pivot_wider(names_from = 'Series Name', values_from = indicator_value)
```


##Pivot df long so dog food types are in their columns again
```{r}
df_wide <- df_long %>% 
  pivot_wider(names_from = dog_food_type, 
              values_from = rankings)
```

## Rename columns

```{r}
# for renaming multiple columns
names(wb_data_tidy) <- c("country", "year", "access_clean_fuels", "access_electricity", "co2_emissions", "fossil_fuel_cons", "water_stress")

# for renaming just 1 column: (takes less time! new name is first in newname( new name = old name))
wb_data_tidy <- wb_data_tidy %>% 
  rename(country_name = country)
```






